\bf Uniformly Ergodic Data-Augmented MCMC for Fitting the General Stochastic Epidemic Model to Incidence Data

### 2022-01-13 Manuscripts/Article/Article.tex
Jasa style

%\usepackage[utf8]{inputenc}
%\usepackage[round,authoryear]{natbib}

\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{url,hyperref}
\usepackage{graphicx}
\usepackage{amsmath,amssymb,array,eucal, amsthm}
\linespread{1.45}
\setlength\parindent{35pt}
\input{../macros}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{color}

@@

\newcommand{\jx}[1]{{\color{blue}{ #1}}}
\newcommand{\ram}[1]{{\color{green}{ #1}}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{lemma}{Lemma}[section]

\begin{document}

	% Keywords command
	\providecommand{\keywords}[1]
	{
		\small	
		\textbf{\textit{Keywords---}} #1
	}

	\title{%Efficient Likelihood-Based Inference for Fitting Stochastic Epidemic Models to Incidence Data via Data Augmentation
		Uniformly Ergodic Data-Augmented MCMC for Fitting the General Stochastic Epidemic Model to Incidence Data}

	\author{
	Rapha\"{e}l Morsomme$^{1}$\footnote{raphael.morsomme@duke.edu} \ and Jason Xu$^{1}$ \\
	\small $^{1}$Department of Statistical Science, Duke University \\
	}
	\date{
	\today
	}	
	\maketitle

	\begin{abstract}
		Stochastic epidemic models provide an interpretable probabilistic description of the spread of a disease through a population. Yet, fitting these models when the epidemic process is only partially observed is a notoriously difficult task due to the intractability of the likelihood for many classical models. To remedy this issue, this article introduces a novel data-augmented MCMC algorithm for fast and exact inference for the stochastic SIR model given discretely observed incidence counts of infection. In a Metropolis-Hastings step, the algorithm jointly proposes the augmented data from a surrogate stochastic process whose dynamics closely resemble those of the SIR process, and from which we can efficiently generate an epidemic that is compatible with the observed data. Not only is the algorithm fast, but since the augmented data are generated from a faithful approximation of the target model, the method can update large portions of the latent space per iteration without prohibitively lowering the acceptance rate. We find that the method explores the high-dimensional latent space efficiently and show that the Markov chain underlying the algorithm is uniformly ergodic.
		The proposed algorithm enables exact Bayesian inference that scales to outbreaks with hundreds of thousands of individuals,  even on a single laptop. We validate its performance via thorough simulation experiments and a case study on the 2013-2015 Ebola outbreak in Western Africa.
	\end{abstract}
	\keywords{Stochastic SIR model; incidence data; exact Bayesian inference; data-augmented MCMC; likelihood-based methods; uniform ergodicity}

	\section{Introduction}
	% Outline: SEM; intractable likelihood; existing methods: approximation (Cauchemez 2008, Fintzi 2020), particle filtering, ABC, data augmentation; poor mixing;

	The efficient control of a disease outbreak requires an understanding of the mechanisms underlying its spread. Mechanistic compartmental models, which describe the transition of individuals between various states, have a long mathematical modeling tradition in epidemiology \cite{Kermack.1927}. Due to their interpretability, they are commonly used to describe the dynamics of an outbreak and typically serve as the main source of information for predicting the course of an outbreak and identifying interventions that could be effective. Originally, deterministic versions of the models were employed by mathematicians and epidemiologists. These models are simple to analyze, but fail to capture the inherent randomness characterizing the spread of a disease. For instance, they cannot be used to estimate the probability of a large-scale outbreak or its expected duration, and do not allow for uncertainty quantification when used within inferential procedures. Stochastic epidemic models (SEM), on the other hand, incorporate the random nature of infections and recoveries and therefore provide more realistic descriptions of the spread of a disease, and in turn more reliable inference from observed data.

	Conducting inference on SEMs is, however, a notably difficult task. Challenges stem from the fact that the observed data typically provide incomplete information on a process that evolves continuously through time, making the likelihood of the model intractable. %Were the infection and recovery times of each individual of the population observed, inference would be straightforward, but this is rarely the case. 
	In practice, one often has access to either incidence data such as weekly counts of new infections, or prevalence data such as the numbers of infectious individuals at discrete reporting times. The marginal likelihood of such partially observed data becomes a computational bottleneck, as it requires a large integration step that accounts for all possible configurations of the missing epidemic process. In particular, direct computation of this likelihood requires the transition probabilities between observation times for which no closed form is available. Moreover, given the size of the transition matrix, classical matrix exponentiation is intractable. %\jx{to do: cite recent papers from Simon Spencer's group too that do so for discrete-time models, and mention classical matrix exponentiation being intractable}.
	Numerical methods to obtain transition probabilities have only recently been developed in \cite{Ho.2018b, Ho.2018}, but the high computational costs limit their approach to moderate sized outbreaks.

@@

	The remainder of the article is structured as follows. Section \eqref{sec:set} provides background information on the inference task. It introduces the stochastic SIR model and explains why conducting inference from partially observed data is a difficult task. Previous works addressing this problem are also presented. Section \eqref{sec:pds} introduces the DA-MCMC algorithm and describes the surrogate process from which the latent data are generated in the Metropolis-Hastings step. The similarity between the surrogate and the target processes is assessed in Section \eqref{sec:qua} and the DA-MCMC is shown to be uniformly ergodic in Section \eqref{sec:uni}. In Section \eqref{sec:sim}, we examine the performance of our algorithm on simulated data and then turn to an analysis of the 2013-2015 Ebola outbreak in Gu\'eck\'edou, Guinea, in Section \eqref{sec:ebo}. Finally, Section \eqref{sec:dis} discusses the findings and concludes the article.

	\section{Background and Prior Work}
	\label{sec:set}
	%Outline: compartmental model; CTMC; rates; likelihood; inference (MLE, gamma conjugacy)

	\subsection{The Stochastic SIR Model}
	\label{sec:sir}
	%Outline: stochastic SIR; X; parameters; transition rate12

	Our point of departure is to consider the stochastic SIR model -- also referred to as the general stochastic epidemic model \cite{Bailey.1975} -- which offers a parsimonious and interpretable representation of the mechanistic dynamics of an epidemic. The stochastic SIR model is a compartmental model in which individuals transition through three states or compartments: susceptible (S), infectious (I) and removed (R). The only possible moves are from S to I (infections) and from I to R (removals). A susceptible individual becomes infected through contact with an infectious individual. Once infected, she is immediately infectious and remains so for some period of time after which she is removed from the process without the possibility of reinfection. In this formulation, demographic dynamics such as births and deaths are ignored since they usually occur at a much slower rate than infections and removals.

	Assuming a closed population of size $n$, the stochastic SIR model consists of a continuous-time vector-valued process
	\begin{equation}
		\label{eq:X}
		\X = \left\lbrace \X(t), t>0\right\rbrace \in \chi_{\X}
	\end{equation}
	with
	\begin{equation}
		\X(t) = \left(X_1(t), \dots, X_n(t)\right) \in \{s, i, r\}^n
	\end{equation}
	where the agent-level subprocess
	$$ X_j(t) = 
	\begin{cases}
		s, & t \in [0, \tau^I_j] \\
		i, & t \in (\tau^I_j, \tau^R_j] \\
		r, & t \in (\tau^R_j, \infty)
	\end{cases}
	,\quad \iton
	$$
	denotes the compartment of individual $j$ at time $t$ with $\tau^I_j$ and $\tau^R_j$ respectively the infection and removal times of individual $j$. If individual $j$ never becomes infectious, we set $\tau^I_j = \tau^R_j = \infty$ and $X_j(t) = s$ for $t \in [0, \infty)$. Since removed individuals do not contribute to the pandemic, it is safe to ignore the individuals removed at time $0$ and write $n = S(0) + I(0)$. Here the space $\chi_{\X}$ denotes the set of trajectories compatible with the evolution of a disease---that is, the set of trajectories in which no infection occurs when the infectious compartment is depleted:
	\begin{equation}
		\label{eq:chi}
		\chi_{\X} = \{\X:\X(t) \in \{s,r\}^n \Rightarrow \X(t+u) = \X(t), \forall u>0 \}.
	\end{equation}

	The stochastic SIR model is specified by the rates at which individuals move from one compartment to another. If we assume a homogeneously mixing population where contacts between individuals occur independently at some constant rate $\beta$, the contacts between two given individuals are said to follow a Poisson process with rate $\beta$. This parameter can be interpreted as the \textit{infection rate}: when a susceptible individual comes into contact with an infectious individual, she immediately becomes infectious.
	If we also make the common assumption that the infectious periods follow independent exponential distributions with rate $\gamma$, then the process \eqref{eq:X} is a time–homogeneous continuous time Markov chain, whose instantaneous transition rates
	are given by the matrix $\Lambda = [\lambda_{\x, \x'}]$ with
	\begin{equation}
	\label{eq:rate}
    	\lambda_{\x, \x'} = 
    	\begin{cases}
    		\beta I(t), & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = s \text{ and } x_j' = i, \\
    		\gamma, & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = i \text{ and } x_j' = r, \\
    		0, & \text{ otherwise.}
    	\end{cases}
	\end{equation}
	where $I(t) = \#\{x_j(t) = i\}$ is the total number of infectious individuals at time $t$. Thus, the individual-level infection and removal rates at time $t$ are respectively $\beta I(t)$ and $\gamma$.

	%Equivalently \jx{The following is not an exact equality, but needs the $o(dt)$ term or similar}
	%	$$
	%	P(\X_t = \x, \X_{t+dt} = \x') =
	%	\begin{cases}
	%		\beta I(t) dt + o(dt), & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = S \text{ and } x_j' = I, \\
	%		\gamma dt + o(dt), & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = I \text{ and } x_j' = R, \\
	%		o(dt), & \text{otherwise}
	%	\end{cases}
	%	$$
	%	where $dt>0$ is small.

@@

	When the Markov process \eqref{eq:X} is completely observed until time $t_{end}$, we obtain the following likelihood \cite{Streftaris.2002}
	\begin{align}
		L(\theta; \X)
		& = \prod_{j \in \mathcal{I}} \beta I(\tau^I_j) \prod_{k \in \mathcal{R}} \gamma \exp\left\lbrace - \int_{0}^{t_{end}}\beta S(t)I(t) + \gamma I(t) dt \right\rbrace  \nonumber \\
		\label{eq:cdl}
		& = \beta^{n_I} \gamma^{n_R}\prod_{j \in \mathcal{I}} I(\tau^I_j) \exp\left\lbrace - \beta \int_{0}^{t_{end}} S(t)I(t) - \gamma \int_{0}^{t_{end}} I(t) dt \right\rbrace
	\end{align}
	describing the complete data trajectory.
	To establish notation,
	$\theta = (\beta, \gamma) \in \chi_{\theta}$ are the model parameters 
	and $\chi_{\theta} = (0, \infty) \times (0, \infty)$ denotes the parameter space, 
	the index sets $\mathcal{I} = \{j: \tau^I_j \in (0, t_{end}]\}$ and $\mathcal{R} = \{j: \tau^R_j \in (0, t_{end}]\}$ denote the individuals that are respectively infected and removed during the observation interval $(0, t_{end}]$,
	$n_I = |\mathcal{I}|$ and $n_R = |\mathcal{R}|$ are the numbers of observed infections and removals,
	and	$S(t) = \#\{X_i(t) = s\}$ is the number of susceptible individuals at time $t$.

	It is straightforward to conduct inference in this continuously observed scenario \cite{Guttorp.2018}. The likelihood \eqref{eq:cdl} belongs to the exponential family, and maximum likelihood estimates of its parameters can be expressed in terms of the sufficient statistics defined above as 
	$$
	\hat{\beta} = \dfrac{n_I}{ \int_{0}^{t_{end}} S(t)I(t)dt}, \qquad \hat{\gamma} = \dfrac{n_R}{ \int_{0}^{t_{end}} I(t)dt}.
	$$
	Note that since the functions $I(t)$ and $S(t)$ are constant between event times, the two integrals correspond to finite sums which are straightforward to compute.
	Furthermore, in a Bayesian context, inference is facilitated by the conjugacy of \eqref{eq:cdl} with the gamma distribution. If we use independent gamma priors
	\begin{equation}
		\label{eq:pri}
		\beta \sim G(a_{\beta}, b_{\beta}), \qquad \gamma \sim G(a_{\gamma}, b_{\gamma})
	\end{equation}
	where $G(a,b)$ denotes the parametrization of the gamma distribution with mean $a/b$ and variance $a/b^2$, we obtain the independent posterior distributions
	\begin{align}
		\label{eq:posterior_theta}
		\beta | \X & \sim G\left( a_{\beta} + n_I , b_{\beta} + \int_{0}^{t_{end}} I(t)S(t) dt\right) \\
		\gamma | \X & \sim G\left( a_{\gamma} + n_R, b_{\gamma} + \int_{0}^{t_{end}} I(t) dt\right)
	\end{align}
	from which one can easily generate independent values to explore the distribution $\pi(\theta|\X)$ via Monte Carlo.

	The likelihood \eqref{eq:cdl} can also be parameterized in terms of $\tilde{\theta} = (\beta, R_0)$ where $R_0 := S(0) \beta / \gamma$ is called the basic reproduction number and corresponds to the expected number of secondary infections caused by an infectious individual in a susceptible population. Using $\tilde{\theta}$ with the independent prior distributions
	\begin{equation}
		\label{eq:pri2}
		\beta \sim G(a_{\beta}, b_{\beta}), \qquad R_0 \sim IG(a_{R}, b_{R}),
	\end{equation}
	where $IG$ denote the inverse-gamma distribution, yields the closed-form full conditional distributions.
	\begin{align}		\label{eq:posterior_theta2}
		\beta | \X, R_0 & \sim G\left( a_{\beta} + n_I + n_R, b_{\beta} + \int_{0}^{t_{end}} S(t)I(t) dt + \frac{S(0)}{R_0} \int_{0}^{t_{end}} I(t) dt\right) \\
		R_0 | \X, \beta & \sim IG\left( a_{R} + n_R, b_{R} + \beta S(0) \int_{0}^{t_{end}} I(t) dt\right)
	\end{align}
	We have observed that the parameterization $\tilde{\theta}$ results in a Markov chain that mixes better and therefore employ it in the simulation studies and data analysis of Section \eqref{sec:per}.

	%	\begin{equation}
%	    \label{eq:lik2}
%	    L(\X|\tilde{\theta}) = \beta^{n_I + n_R} R_0^{-n_R} S(0)^{n_R} \prod_{j\in\mathcal{I}} I(\tau^I_j) \exp\left\lbrace - \beta \int_{0}^{t_{end}} I(t)S(t) - \frac{\beta S(0)}{R_0} \int_{0}^{t_{end}} I(t) dt \right\rbrace,
%	\end{equation}

@@

	In practice, inference is complicated by the fact that the epidemic process \eqref{eq:X} is only partially observed. When we do not observe all event times, we do not have access to the sufficient statistics needed to evaluate the likelihood \eqref{eq:cdl}. Various types of partially observed data typify real data and have been considered in the literature, such as observing only the removal times \cite{Gibson.1998, ONeill.1999} or the number of infectious individuals at discrete points in time \cite{Fintzi.2017}. For instance, the former arises in animal experiments in which positive cases are immediately isolated from the rest of the population and no longer contribute to disease spread. In such cases, exact times of removals are known, but this information is unavailable in a typical observational study.

	In this article, we focus our attention on partial data consisting of \textit{incidence} counts of infections during given time intervals, e.g.\ weekly infection counts. Such data arise when an infectious individual is identified some time after onset and may continue to infect others after being tested positive for the disease. Given an observation schedule $t_{0:K}$ ($K \ge 1$) with $0 = t_0 < t_1 < \dots < t_K = t_{end}$, the observed data consist of the $K$-dimensional vector $\Y = I_{1:K}$ where $I_k = \#\{\tau^I_j \in (t_{k-1}, t_k]\}$ is the number of infections during the $k^{\text{th}}$ interval.

	This form of observed data is motivated by the $2013$–$2016$ outbreak of the Ebola virus, the largest outbreak since the discovery of the virus in 1976, which resulted in at least 11,000 deaths, mainly in three Western African countries (Guinea, Liberia and Sierra Leone). The data available for this outbreak consist of weekly number of positive tests. This outbreak was characterized by the large number of infections that occurred after individuals were tested positive for the virus, giving rise to noisy incidence data. Numerous infections took place at the hospitals where infected individuals received treatment as well as at the funerals of people deceased from the virus\footnote{In most regions impacted by the Ebola virus, touching the body of the deceased during funerals is a tradition. Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.} \cite{Coltart.2017}. We therefore decide to model the weekly number of positive tests as the weekly number of infections in a SIR model.

@@

	In a Bayesian framework, the posterior distribution of the parameters given the observed data is formally related to the complete data likelihood \eqref{eq:cdl} via integration:
	\begin{equation}
		\label{eq:pdl}
		\pi(\theta|\Y) 
		\propto \pi(\theta) L(\theta; \Y) = \pi(\theta) \int_{\chi_\x} L(\theta; \x) \delta_{\Y}(\x) d\x
		%\mathbf{1}\{\Y \overset{\Delta}{=} \x\}
	\end{equation}
	where $\pi(\theta)$ is the prior distribution on $\theta$, and
	$\delta_{\Y}(\x) = 1$ if $\x$ is compatible with the observed infection counts $\Y$ and $0$ otherwise; that is, the partial data likelihood $L(\theta; \Y)$ consists of a high-dimensional integral over all epidemic paths that agree with the observed data. This marginalization step has no known closed-form solution and presents computational challenges even for a population of moderate size.

>>>>

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{subcaption}
\usepackage{array}
\usepackage{amsthm}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

\newcommand{\jx}[1]{{\color{blue}{ #1}}}
\newcommand{\ram}[1]{{\color{green}{ #1}}}
\begin{document}

	%\bibliographystyle{natbib}

	\def\spacingset#1{\renewcommand{\baselinestretch}%
		{#1}\small\normalsize} \spacingset{1}

	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

	\if1\blind
	{
		\title{\bf Uniformly Ergodic Data-Augmented MCMC for Fitting the General Stochastic Epidemic Model to Incidence Data}
		\author{Rapha\"{e}l Morsomme\thanks{
				The authors gratefully acknowledge \textit{please remember to list all relevant funding sources in the unblinded version}}\hspace{.2cm}\\
			Department of Statistical Science, University of Duke\\
			and \\
			Jason Xu \\
			Department of Statistical Science, University of Duke}
		\maketitle
	} \fi

	\if0\blind
	{
		\bigskip
		\bigskip
		\bigskip
		\begin{center}
			{\LARGE\bf Uniformly Ergodic Data-Augmented MCMC for Fitting the General Stochastic Epidemic Model to Incidence Data}
		\end{center}
		\medskip
	} \fi

	\bigskip
	\begin{abstract}
		Stochastic epidemic models provide an interpretable probabilistic description of the spread of a disease through a population. Yet, fitting these models when the epidemic process is only partially observed is a notoriously difficult task due to the intractability of the likelihood for many classical models. To remedy this issue, this article introduces a novel data-augmented MCMC algorithm for fast and exact Bayesian inference for the stochastic SIR model given discretely observed infection incidence counts. In a Metropolis-Hastings step, new event times of the latent data are jointly proposed from a surrogate process that closely resembles the SIR, and from which we can efficiently generate epidemics compatible with the observed data.

		The proposed DA-MCMC algorithm is fast and, since the latent data are generated from a faithful approximation of the target model, a large portion thereof can be updated per iteration without prohibitively lowering the acceptance rate. We find that the method explores the high-dimensional latent space efficiently and scales to outbreaks with hundreds of thousands of individuals, and we show that the Markov chain underlying the algorithm is uniformly ergodic. We validate its performance via thorough simulation experiments and a case study on the 2013-2015 Ebola outbreak in Western Africa.
	\end{abstract}

	\noindent%
	{\it Keywords:}  Stochastic epidemic model; incidence data; exact Bayesian inference; likelihood-based inference
	\vfill

	\newpage
	\spacingset{1.9} % DON'T change the spacing!

\section{Introduction}
\label{sec:intro}

The efficient control of a disease outbreak requires an understanding of the mechanisms underlying its spread. Mechanistic compartmental models, which describe the transition of individuals between various states, have a long mathematical modeling tradition in epidemiology dating back to \cite{Kermack.1927}. Due to their interpretability, they are commonly used to describe the dynamics of an outbreak and typically serve as the main source of information for predicting the course of an outbreak and identifying interventions that could be effective. Originally, deterministic versions of the models were employed by mathematicians and epidemiologists. These models are simple to analyze, but fail to capture the inherent randomness characterizing the spread of a disease. For instance, they cannot be used to estimate the probability of a large-scale outbreak or its expected duration, and do not allow for uncertainty quantification when used within inferential procedures. Stochastic epidemic models (SEM), on the other hand, incorporate the random nature of infections and recoveries and therefore provide more realistic descriptions of the spread of a disease, and in turn more reliable inference from observed data.

Conducting inference on SEMs is, however, a notably difficult task. Challenges stem from the fact that the observed data typically provide incomplete information on a process that evolves continuously through time, making the likelihood of the model intractable. %Were the infection and recovery times of each individual of the population observed, inference would be straightforward, but this is rarely the case.
In practice, one often has access to either incidence data such as weekly counts of new infections, or prevalence data such as the numbers of infectious individuals at discrete reporting times. The marginal likelihood of such partially observed data becomes a computational bottleneck, as it requires a large integration step that accounts for all possible configurations of the missing epidemic process. In particular, direct computation of this likelihood requires the transition probabilities between observation times for which no closed form is available. Moreover, given the size of the transition matrix, classical matrix exponentiation is intractable.
Numerical methods to obtain transition probabilities have recently been developed in \cite{Ho.2018}, but the high computational costs limit their approach to moderate sized outbreaks.

@@

The remainder of the article is structured as follows. Section \ref{sec:set} provides background information on the inference task. It introduces the stochastic SIR model and explains why conducting inference from partially observed data is a difficult task. Previous works addressing this problem are also presented. Section \ref{sec:pds} introduces the DA-MCMC algorithm and describes the surrogate process from which the latent data are generated in the Metropolis-Hastings step. The similarity between the surrogate and the target processes is assessed in Section \ref{sec:qua} and the DA-MCMC is shown to be uniformly ergodic in Section \ref{sec:uni}. In Section \ref{sec:sim}, we examine the performance of our algorithm on simulated data and then turn to an analysis of the 2013-2015 Ebola outbreak in Gu\'eck\'edou, Guinea, in Section \ref{sec:ebo}. Finally, Section \ref{sec:dis} discusses the findings and concludes the article.

\section{Background and Prior Work}
\label{sec:set}

\subsection{The Stochastic SIR Model}
\label{sec:sir}
Our point of departure is to consider the stochastic SIR model introduced by \cite{Bailey.1975} -- also referred to as the general stochastic epidemic model -- which offers a parsimonious and interpretable representation of the mechanistic dynamics of an epidemic. The stochastic SIR model is a compartmental model in which individuals transition through three states or compartments: susceptible (S), infectious (I) and removed (R). The only possible moves are from S to I (infections) and from I to R (removals). A susceptible individual becomes infected through contact with an infectious individual. Once infected, she is immediately infectious and remains so for some period of time after which she is removed from the process without the possibility of reinfection. In this formulation, demographic dynamics such as births and deaths are ignored since they usually occur at a much slower rate than infections and removals.

Assuming a closed population of size $n$, the stochastic SIR model consists of a continuous-time vector-valued process
\begin{equation}
	\label{eq:X}
	\mathbf{X} = \left\lbrace \mathbf{X}(t), t>0\right\rbrace \in \chi_{\mathbf{X}}
\end{equation}
with
\begin{equation}
	\mathbf{X}(t) = \left(X_1(t), \dots, X_n(t)\right) \in \{s, i, r\}^n
\end{equation}
where the agent-level subprocess
$$ X_j(t) = 
\begin{cases}
	s, & t \in [0, \tau^I_j] \\
	i, & t \in (\tau^I_j, \tau^R_j] \\
	r, & t \in (\tau^R_j, \infty)
\end{cases}
,\quad i = 0, \dots, n
$$
denotes the compartment of individual $j$ at time $t$ with $\tau^I_j$ and $\tau^R_j$ respectively the infection and removal times of individual $j$. If individual $j$ never becomes infectious, we set $\tau^I_j = \tau^R_j = \infty$ and $X_j(t) = s$ for $t \in [0, \infty)$. Since removed individuals do not contribute to the pandemic, it is safe to ignore the individuals removed at time $0$ and write $n = S(0) + I(0)$. Here the space $\chi_{\mathbf{X}}$ denotes the set of trajectories compatible with the evolution of a disease---that is, the set of trajectories in which no infection occurs when the infectious compartment is depleted:
\begin{equation}
	\label{eq:chi}
	\chi_{\mathbf{X}} = \{\mathbf{X}:\mathbf{X}(t) \in \{s,r\}^n \Rightarrow \mathbf{X}(t+u) = \mathbf{X}(t), \forall u>0 \}.
\end{equation}

The stochastic SIR model is specified by the rates at which individuals move from one compartment to another. If we assume a homogeneously mixing population where contacts between individuals occur independently at some constant rate $\beta$, the contacts between two given individuals are said to follow a Poisson process with rate $\beta$. This parameter can be interpreted as the \textit{infection rate}: when a susceptible individual comes into contact with an infectious individual, she immediately becomes infectious.
If we also make the common assumption that the infectious periods follow independent exponential distributions with rate $\gamma$, then the process \eqref{eq:X} is a time?homogeneous continuous time Markov chain, whose instantaneous transition rates
are given by the matrix $\Lambda = [\lambda_{\mathbf{x}, \mathbf{x}'}]$ with
\begin{equation}
	\label{eq:rate}
	\lambda_{\mathbf{x}, \mathbf{x}'} = 
	\begin{cases}
		\beta I(t), & \mathbf{x} \text{ and } \mathbf{x}' \text{ only differ at position } j \text{ with } x_j = s \text{ and } x_j' = i, \\
		\gamma, & \mathbf{x} \text{ and } \mathbf{x}' \text{ only differ at position } j \text{ with } x_j = i \text{ and } x_j' = r, \\
		0, & \text{ otherwise.}
	\end{cases}
\end{equation}
where $I(t) = \#\{x_j(t) = i\}$ is the total number of infectious individuals at time $t$. Thus, the individual-level infection and removal rates at time $t$ are respectively $\beta I(t)$ and $\gamma$.

@@

When the Markov process \eqref{eq:X} is completely observed until time $t_{end}$, we obtain the following likelihood \citep{Streftaris.2002}
\begin{align}
	L(\theta; \mathbf{X})
	& = \prod_{j \in \mathcal{I}} \beta I(\tau^I_j) \prod_{k \in \mathcal{R}} \gamma \exp\left\lbrace - \int_{0}^{t_{end}}\beta S(t)I(t) + \gamma I(t) dt \right\rbrace  \nonumber \\
	\label{eq:cdl}
	& = \beta^{n_I} \gamma^{n_R}\prod_{j \in \mathcal{I}} I(\tau^I_j) \exp\left\lbrace - \beta \int_{0}^{t_{end}} S(t)I(t) - \gamma \int_{0}^{t_{end}} I(t) dt \right\rbrace
\end{align}
describing the complete data trajectory.
To establish notation,
$\theta = (\beta, \gamma) \in \chi_{\theta}$ are the model parameters 
and $\chi_{\theta} = (0, \infty) \times (0, \infty)$ denotes the parameter space, 
the index sets $\mathcal{I} = \{j: \tau^I_j \in (0, t_{end}]\}$ and $\mathcal{R} = \{j: \tau^R_j \in (0, t_{end}]\}$ denote the individuals that are respectively infected and removed during the observation interval $(0, t_{end}]$,
$n_I = |\mathcal{I}|$ and $n_R = |\mathcal{R}|$ are the numbers of observed infections and removals,
and	$S(t) = \#\{X_i(t) = s\}$ is the number of susceptible individuals at time $t$.

It is straightforward to conduct inference in this continuously observed scenario %\cite{Guttorp.2018}
. The likelihood \eqref{eq:cdl} belongs to the exponential family, and maximum likelihood estimates of its parameters can be expressed in terms of the sufficient statistics defined above as 
$$
\hat{\beta} = \dfrac{n_I}{ \int_{0}^{t_{end}} S(t)I(t)dt}, \qquad \hat{\gamma} = \dfrac{n_R}{ \int_{0}^{t_{end}} I(t)dt}.
$$
Note that since the functions $I(t)$ and $S(t)$ are constant between event times, the two integrals correspond to finite sums which are straightforward to compute.
Furthermore, in a Bayesian context, inference is facilitated by the conjugacy of \eqref{eq:cdl} with the gamma distribution. If we use independent gamma priors
\begin{equation}
	\label{eq:pri}
	\beta \sim G(a_{\beta}, b_{\beta}), \qquad \gamma \sim G(a_{\gamma}, b_{\gamma})
\end{equation}
where $G(a,b)$ denotes the parametrization of the gamma distribution with mean $a/b$ and variance $a/b^2$, we obtain the independent posterior distributions
\begin{align}
	\label{eq:posterior_theta}
	\beta | \mathbf{X} & \sim G\left( a_{\beta} + n_I , b_{\beta} + \int_{0}^{t_{end}} I(t)S(t) dt\right) \\
	\gamma | \mathbf{X} & \sim G\left( a_{\gamma} + n_R, b_{\gamma} + \int_{0}^{t_{end}} I(t) dt\right)
\end{align}
from which one can easily generate independent values to explore the distribution $\pi(\theta|\mathbf{X})$ via Monte Carlo.

The likelihood \eqref{eq:cdl} can also be parameterized in terms of $\tilde{\theta} = (\beta, R_0)$ where $R_0 := S(0) \beta / \gamma$ is called the basic reproduction number and corresponds to the expected number of secondary infections caused by an infectious individual in a susceptible population. Using $\tilde{\theta}$ with the independent prior distributions
\begin{equation}
	\label{eq:pri2}
	\beta \sim G(a_{\beta}, b_{\beta}), \qquad R_0 \sim IG(a_{R}, b_{R}),
\end{equation}
where $IG$ denote the inverse-gamma distribution, yields the closed-form full conditional distributions.
\begin{align}		\label{eq:posterior_theta2}
	\beta | \mathbf{X}, R_0 & \sim G\left( a_{\beta} + n_I + n_R, b_{\beta} + \int_{0}^{t_{end}} S(t)I(t) dt + \frac{S(0)}{R_0} \int_{0}^{t_{end}} I(t) dt\right) \\
	R_0 | \mathbf{X}, \beta & \sim IG\left( a_{R} + n_R, b_{R} + \beta S(0) \int_{0}^{t_{end}} I(t) dt\right)
\end{align}
We have observed that the parameterization $\tilde{\theta}$ results in a Markov chain that mixes better and therefore employ it in the simulation studies and data analysis of Section \ref{sec:per}.

@@

In practice, inference is complicated by the fact that the epidemic process \eqref{eq:X} is only partially observed. When some event times are unobserved, we do not have access to the sufficient statistics needed to evaluate the likelihood \eqref{eq:cdl}. Various types of partially observed data typify real data and have been considered in the literature. These include the removal times in \cite{Gibson.1998} and \cite{ONeill.1999} or the number of infectious individuals at discrete points in time in \cite{Fintzi.2017}. For instance, the former arises in animal experiments in which positive cases are immediately isolated from the rest of the population and no longer contribute to disease spread. In such cases, exact times of removals are known, but this information is unavailable in a typical observational study.

In this article, we focus our attention on partial data consisting of \textit{incidence} counts of infections during given time intervals, e.g.\ weekly infection counts. Such data arise when an infectious individual is identified some time after onset and may continue to infect others after being tested positive for the disease. Given an observation schedule $t_{0:K}$ ($K \ge 1$) with $0 = t_0 < t_1 < \dots < t_K = t_{end}$, the observed data consist of the $K$-dimensional vector $\mathbf{Y} = I_{1:K}$ where $I_k = \#\{\tau^I_j \in (t_{k-1}, t_k]\}$ is the number of infections during the $k^{\text{th}}$ interval.

This form of observed data is motivated by the $2013$-$2015$ outbreak of the Ebola virus, the largest outbreak since the discovery of the virus in 1976, which resulted in at least 11,000 deaths, the majority of which occurred in three Western African countries (Guinea, Liberia and Sierra Leone). The data available for this outbreak consist of weekly number of positive tests. This outbreak was characterized by the large number of infections that occurred after individuals were tested positive for the virus, giving rise to noisy incidence data. \cite{Coltart.2017} reports that numerous infections took place at the hospitals where infected individuals received treatment as well as at the funerals of people deceased from the virus\footnote{In most regions impacted by the Ebola virus, touching the body of the deceased during funerals is a tradition. Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.}. We therefore model the weekly number of positive tests as the weekly number of infections in a SIR model.

@@

In a Bayesian framework, the posterior distribution of the parameters given the observed data is formally related to the complete data likelihood \eqref{eq:cdl} via integration:
\begin{equation}
	\label{eq:pdl}
	\pi(\theta|\mathbf{Y}) 
	\propto \pi(\theta) L(\theta; \mathbf{Y}) = \pi(\theta) \int_{\chi_\mathbf{x}} L(\theta; \mathbf{x}) \delta_{\mathbf{Y}}(\mathbf{x}) d\mathbf{x}
	%\mathbf{1}\{\Y \overset{\Delta}{=} \mathbf{x}\}
\end{equation}
where $\pi(\theta)$ is the prior distribution on $\theta$, and
$\delta_{\mathbf{Y}}(\mathbf{x}) = 1$ if $\mathbf{x}$ is compatible with the observed infection counts $\mathbf{Y}$ and $0$ otherwise; that is, the partial data likelihood $L(\theta; \mathbf{Y})$ consists of a high-dimensional integral over all epidemic paths that agree with the observed data. This marginalization step has no known closed-form solution and presents computational challenges even for a population of moderate size.
